---
title: "MATH 629 HW 4 Solutions"
author: "Drew Lazar"
output:
  pdf_document:
    keep_tex: yes
  html_document:
    df_print: paged
---
```{r results='hide', include=FALSE, message=FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy=TRUE)
```

```{r setup,results='hide',include=FALSE,message=FALSE,warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/GitStuff/Survival_Analysis/MATH_629/Data")
```

## Cleaning up and loading necessary packages 
```{r}
rm(list=ls())
library(survival)
library(dplyr)
```


## 1 loading the data 
```{r}
Ven.reset <-read.csv("VenresetMft.csv", header = TRUE)
```
## 2 Test the PH assumption
## 2i. GOF using Schoenfeld Residuals 
```{r}
Y<-Surv(Ven.reset$eventtime,Ven.reset$status)
Coxph.addicts=coxph(Y~Setting+LO2+Mft,data=Ven.reset)
summary(Coxph.addicts)
cox.zph(Coxph.addicts,transform=rank)
```
## 2ii. Testing MFt with Setting and LO2 in the model 
```{r}
#Stratify Ventilator Data set by TR
Ven.reset0<-Ven.reset[Ven.reset$Mft==0, ]
Ven.reset1<-Ven.reset[Ven.reset$Mft==1, ]
#Create Survival Objects for both strata 
Y0<-Surv(Ven.reset0$eventtime,Ven.reset0$status==1)
Y1<-Surv(Ven.reset1$eventtime,Ven.reset1$status==1)
#Fit Cox PH models to both strata
Coxph.Ven.m0=coxph(Y0~LO2+Setting,data=Ven.reset0)
Coxph.Ven.m1=coxph(Y1~LO2+Setting,data=Ven.reset1)
#get the overall mean of LO2 and Setting 
meanLO2=mean(Ven.reset$LO2)
meanSetting=mean(Ven.reset$Setting)
#plot our adjusted survival curves 
pattern=data.frame(LO2=meanLO2,Setting=meanSetting)
plot(survfit(Coxph.Ven.m0,newdata=pattern),fun="cloglog",conf.int=F,xlim=c(1,23),ylim=c(-3.9,2.3),main="Adjusted log-log for Mft=0 vs Mft=1",col=c('blue'))
par(new=TRUE)
plot(survfit(Coxph.Ven.m1,newdata=pattern),fun="cloglog",conf.int=F,xlim=c(1,23),ylim=c(-3.9,2.3),col=c('red'))
legend("bottomright",cex=1.5,c("Mft=0","Mft=1"),lty=c("solid"),col=c("blue","red"))
```
**Comment:**The Schoenfeld Residuals for Setting, LO2 and Mft are 0.1658, 0.9808 and 0.0017, respectively.  Also, the adjusted log-log plots for Mft clearly cross and are not parallel throughout the study. Both of these suggest that Mft doesn't satisfy the PH assumption with Setting and LO2 in the model.

## 3. Stratification by Mft
## 3i. Fit a stratified model without interaction. 
```{r}
#i Fit a stratified model 
Y<-Surv(Ven.reset$eventtime,Ven.reset$status==1)
coxph.Ven.m1<-coxph(Y~Setting + LO2 + strata(Mft),data=Ven.reset)
summary(coxph.Ven.m1)
```
**Comment:**Our fitted stratified Cox PH model without interaction is: $h_g(X,t)=h_{g0}(t)\exp(0.96368X_1+0.83814X_2)$ for $g=1,2$ with $X_1=Setting$ and $X_2=LO2$.

## 3ii. Wald test to test for significance of Setting and LO2 
**Comment:** Both setting an LO2 have very small p-values (5.14e-14 and <2e-16, respectively) and thus are significant at $\alpha=0.05$.

## 3iii. Likelihood ratio test to test for overall significance of Setting and LO2. 
```{r}
coxph.Ven.m1$loglik
CSstat = -2*(coxph.Ven.m1$loglik[1]-coxph.Ven.m1$loglik[2])
print(paste("The value of our test statistic is", CSstat))
CV = qchisq(.95,df=2)
print(paste("Our critical value is:",CV))
CSstat>CV
pvalue1=pchisq(CSstat,df = 2,lower.tail = FALSE)
print(paste("Our p value is:",pvalue1))
```
**Comment:**Our test statistic has a value of 163.6716 and our critical value is 5.9915. As 163.6716>5.9915 we reject $H_0: \beta_1=\beta_2=0$ in $h_g(X,t)=h_{g0}(t)\exp(\beta_1 X_1+ \beta_2 X_2)$.

## 3iv. 
**Comment:**For every unit increase in Setting the hazard increases by a factor of $e^{0.93998}=2.560$ and for every unit increase in LO2 the hazard increases by a factor of $e^{0.84471}=2.327$. As we assume no interaction, this is regardless of the level of Mft.    

## 3v. Plotting adjusted Survival Curves 
```{r}
meanLO2=mean(Ven.reset$LO2)
pattern1=data.frame(Setting=0,LO2=meanLO2)
plot(survfit(coxph.Ven.m1,newdata=pattern1),col=c('blue','red'),conf.int=F,main="Adjusted survival for Setting=0, 1 and 2, mean(LO2) from stratified Cox PH model")
pattern2=data.frame(Setting=1,LO2=meanLO2)
par(new=TRUE)
plot(survfit(coxph.Ven.m1,newdata=pattern2),col=c('blue','red'), lty=c('dashed'),conf.int=F)
pattern3=data.frame(Setting=2,LO2=meanLO2)
par(new=TRUE)
plot(survfit(coxph.Ven.m1,newdata=pattern3),col=c('blue','red'), lty=c('dotted'),conf.int=F)
legend("topright",cex=.75,c("Setting=0, Mft=0","Setting=0, Mft=1","Setting=1, Mft=0","Setting=1,Mft=1","Setting=2, Mft=0","Setting=2, Mft=1"),lty=c("solid","solid","dashed","dashed","dotted","dotted"),col=c("blue","red"))
```
**Comment:**The curves do not suggest significant interaction between Setting and Mft adjusted for LO2. The changes in Survival experience from Setting=0 to Setting=1 to Setting=2 looks quite similar for Mft=0 and Mft=1. 

## 4. Fitting and examining a stratified Cox PH model with interaction
## 4i. 
```{r}
coxph.Ven.int.m1<-coxph(Y~Setting+LO2+Mft:LO2+Mft:Setting+strata(Mft),data=Ven.reset)
summary(coxph.Ven.int.m1)
```
**Comment:**Our fitted stratified Cox PH model with interaction is: 
\[
h_g(X,t)=h_{g0}(t)\exp(0.85827X_1+0.80253X_2 + 0.11042X_1*X_3 +0.16728X_2*X_3)
\]
for $g=1,2$ with $X_1=Setting$, $X_2=LO2$ and $X_3=Mft$.

## 4ii. 
```{r}
CSstat = -2*(coxph.Ven.m1$loglik[2]-coxph.Ven.int.m1$loglik[2])
print(paste("The value of the test statistic is:",CSstat))
CV = qchisq(.95,df=2)
print(paste("The critical value is:",CV))
CSstat>CV
pvalue2=pchisq(CSstat, 2, lower.tail = FALSE)
print(paste("The p value is:",pvalue2))
```
**Comment:**Our test statistic has a value of 0.6511 and our critical value is 5.9915. As 0.6511<5.9915 we fail to reject $H_0: \delta_1=\delta_2=0$ in 
\[
h_g(X,t)=h_{g0}(t)\exp(\beta_1 X_1+ \beta_2 X_2 + \delta_1 X_1*X_3 + \delta_2 X_3*X_3)
\]

## 4iii. 
```{r}
#HR for Setting and 95% CI for HR for Setting with MFt=0. 
b1=coxph.Ven.int.m1$coefficients[1]
exp(b1)
exp(b1-1.96*0.18159)
exp(b1+1.96*0.18159)
```
```{r}
#We use a trick here and recode Mft=0 as Mft2=-1 and Mft=1 as Mft2=0 and refit the model with Mft2 rather than Mft. 
Ven.reset$Mft2=Ven.reset$Mft-1
coxph.Ven.int.m2<-coxph(Y~Setting+LO2+Mft2:LO2+Mft2:Setting+strata(Mft2),data=Ven.reset)
summary(coxph.Ven.int.m2)
#HR for Setting and 95% CI for HR for Setting with MFt=1. 
b1=coxph.Ven.int.m2$coefficients[1]
exp(b1)
exp(b1-1.96*0.1779)
exp(b1+1.96*0.1779)
```
**Comment:** Mft=0: Estimated HR for Setting is 2.359079 and 95% CI:(1.652609, 3.367558). \
Mft=1: Estimated HR for Setting is 2.788632 and 95% CI: (1.967703,3.952052).  \
The HRs are similar and the CIs overlap which suggests there is not a significant interaction effect. \
This agrees with the interaction term with Setting having a p-value of 0.501 (using Wald test) which does not suggest significant interaction. 

## 5.   
The model is
\begin{align*}
h(t;X)=h_0(t)\exp(&\beta_1 X_4 +  \beta_2 X_5 + \beta_3 X_6 + \beta_4 X_4*X_1 + \beta_5 X_5*X_1 + \beta_6 X_6*X_1 \\
                  &+\beta_4 X_4*X_1 + \beta_5 X_5*X_1 + \beta_6 X_6*X_1)
\end{align*}
\begin{enumerate}
\item If the patient Nasopharynx throat cancer $X_1=1$ and $X_2=0$. Let 
\begin{align*}
X &= (X_1,X_2,X_4,X_5,X_6) = (1,0,X_4,X_5,1) \\
\tilde{X} &= (\tilde{X}_1,\tilde{X}_2,\tilde{X}_4,\tilde{X}_5,\tilde{X}_6) = (1,0,X_4,X_5,0)
\end{align*} 
The estimated HR for chemotherapy vs. surgery ($X_6=1$ vs. $X_6=0$) is
\[
\hat{HR} = \frac{\hat{h}(t;X)}{\hat{h}(t;\tilde{X})} = \hat{\beta}_3 + \hat{\beta}_6 = \exp(0.45671) + \exp(-0.307024) = 2.3145
\]
\item Similarly, if the patient has Oropharynx throat cancer $X_1=0$ and $X_2=1$ and the estimated HR for chemotherapy vs. surgery ($X_6=1$ vs. $X_6=0$) is
\[
\hat{HR} = \hat{\beta}_3 + \hat{\beta}_9 = \exp(0.45671) + \exp(0.907024) = 4.05581
\]
\item If the patient has Oropharynx throat cancer $X_1=0$ and $X_2=0$ and the estimated HR for chemotherapy vs. surgery ($X_6=1$ vs. $X_6=0$) is
\[
\hat{HR} = \hat{\beta}_3 = \exp(0.45671) = 1.57887.
\]
\end{enumerate}
Note: A patient's age and BMI have no effect on the estimated HR as neither interaction between age and $X6$ nor interaction between BMI and $X6$ are included in the model.  
## 6. 
```{r}
#Create and plot our partial likelihood. 
f1.1<-function(b) 1/(exp(b*2)+2*exp(1.5*b)+1)
f1.2<-function(b) exp(2*b)/(exp(2*b)+2*exp(1.5*b))
f2.1<-function(b) exp(b)/(2*exp(b)+exp(2*b))
f2.2<-function(b) exp(b)/(exp(b)+exp(2*b))
f<-function(b) f1.1(b)*f1.2(b)*f2.1(b)*f2.2(b)
x <- seq(-20,10,0.01)
plot(x, f(x))   
```
```{r}
#optimize our function to get an estimate. We minizimize the negative of f (g) to maximize f. 
g=function(b) -f(b) 
optimize(g, lower = -5, upper = 0)
```
```{r}
time<-c(2,3,5,6,6,7,8);status<-c(1,1,0,1,0,1,1);Dose<-c(0,2,1,1,1.5,2,1.5);
Type<-c(0,0,1,1,0,1,0)
Alcohol=data.frame(time,status,Dose,Type)
S<-Surv(Alcohol$time,Alcohol$status==1)
coxph(S~Dose+strata(Type),data=Alcohol)
```
**Comment:** Our model is: 
\[
h_g(t;Coupon) = h_{0g}(t)\exp(\beta * Coupon) \text{ for } g=1,2
\]
where $g=1,2$ indicates two strata of medication type. Maximizing our partial likelihood returns an estimate $\hat{\beta}=-1.844587$ and using the coxph routine in R return an estimate of -1.6913. Using R to estimate the effect of a unit increase in Dose on the hazard (i.e. the HR) we get $\hat{HR}=\exp(-1.6913)=0.1843$. That is, for every unit increase in Dose, the hazard goes down by 1-0.1843=0.8157=81.57\% regardless of type (as there is no interaction).  
## Chapter 6
## 6 i. 
```{r}
#Our adjusted survival Curves. 
Coxph.ven3=coxph(Y~Setting+LO2+Mft,data=Ven.reset)
settingmean=mean(Ven.reset$Setting); LO2mean=mean(Ven.reset$LO2); 
pattern1=data.frame(Setting=settingmean,LO2=LO2mean,Mft=0);
pattern2=data.frame(Setting=settingmean,LO2=LO2mean,Mft=1);
plot(survfit(Coxph.ven3,newdata=pattern1),conf.int=F,main="Adjusted survival for Mft=0 vs Mft=1, mean(Setting), mean(LO2)",col=c("blue"))
par(new=TRUE)
plot(survfit(Coxph.ven3,newdata=pattern2),conf.int=F,col=c("red"))
legend("topright",c("Mft=0","Mft=1"), lty=c("solid"),col=c('blue','red'))
```
##6 ii.
```{r}
#Based on the adjusted survival curve we choose t=4 as a split (or cut) point. 
Venreset.cp4=survSplit(Ven.reset,cut=4,end="eventtime",event="status",start="start",id="id")
Venreset.cp4$hv2=Venreset.cp4$Mft*(Venreset.cp4$start>=4)
Y4=Surv(Venreset.cp4$start,Venreset.cp4$eventtime,Venreset.cp4$status)
coxph.Venreset.hs2<-coxph(Y4 ~ Setting + LO2 + Mft + hv2 + cluster(id),data=Venreset.cp4)
summary(coxph.Venreset.hs2)
```
**Comment:**Our model is 
\[
h(t;X)=h(t)\exp(\beta_1 Setting + \beta_2 LO2 + \delta h(t)*Mft )
\]
where 
\[
h(t) = \begin{cases} 1 \text{ if } t \ge 0 \\
                     0 \text{ if } t < 0.
        \end{cases}
\]
As $\delta$ has p-value of $0.0869$ the interaction term is significant at level $\alpha=0.05$. 
The estimated HR is 
\[
\hat{HR}=\exp(\hat{\beta}_1 + \hat{\delta} h(t))
\]

\begin{enumerate}
\item[]The estimated HR before time $t=4$ is $\hat{HR}=\exp(1.01996)=2.77308$. 
\item[] The estimated hazard ratio after time $t=4$ is $\hat{HR}=\exp(1.01996+0.63812)=5.24922.$
\end{enumerate}

## 6iii. 
```{r}
#We fit an extended model with h(t)=ln(t)
Venreset.cp=survSplit(Ven.reset,cut=Ven.reset$eventtime[Ven.reset$status==1],end="eventtime",event="status",start="start",id="id")
Venreset.cp$t.Mft=Venreset.cp$Mft*log(Venreset.cp$eventtime)
#Create an extended survival object
YE<-Surv(Venreset.cp$start,Venreset.cp$eventtime,Venreset.cp$status)
#Test for clinic
coxph.venreset.ct<-coxph(YE ~ LO2 + Setting + Mft + t.Mft + cluster(id),data=Venreset.cp)
summary(coxph.venreset.ct)
```
**Comment:**Our model is 
\[
h(t;X)=h(t)\exp(\beta_1 Setting + \beta_2 LO2 + \delta \ln(t)*Mft )
\]
As $\delta$ has p-value of $0.000282$ the interaction term is significant at level $\alpha=0.05$. 
The estimated HR is 
\[
\hat{HR}=\exp(\hat{\beta}_1 + \hat{\delta} \ln(t))
\]
\begin{enumerate}
\item[] At time $t=1.9$ we have $\hat{HR}=\exp(1.01996+0.49864ln(1.9))=3.8191.$.
\item[] At time $t=4.5$ we have $\hat{HR}\exp(1.01996+0.49864ln(4.5))=5.87058.$
\item[] At time $t=10$ we have $\hat{HR}\exp(1.01996+0.49864ln(4.5))=8.74184.$
\end{enumerate}
Notice that $\hat{HR}$ at time 1.9 and 4.5 are similar to ones provided by our Heaviside model (before time t=4 and after time t=4). 